---
title: "Olive_Oil"
author: "Jean Dos Santos"
date: "April 2020"
output:
  html_document:
    highlight: tango
    number_sections: yes
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = FALSE)
```

# Importing Packages

```{r}
# install.packages("pacman", quiet = TRUE, verbose = FALSE)
library(pacman, quietly = TRUE, verbose = FALSE)

pacman::p_load(cepp, plyr,caret, tidyverse, tidyselect, janitor, gdata, tools, mclust, install = TRUE, update = FALSE)
```

This data is from a paper by Forina, Armanino, Lanteri, Tiscornia (1983) Classification of Olive Oils from their Fatty Acid Composition, in Martens and Russwurm (ed) Food Research and Data Anlysis. We thank Prof. Michele Forina, University of Genova, Italy for making this dataset available.

https://rdrr.io/cran/cepp/man/data-olive.html

`region` Three super-classes of Italy: North, South and the island of Sardinia

`area` Nine collection areas: three from North, four from South and 2 from Sardinia

palmitic, palmitoleic, stearic, oleic, linoleic, linolenic, arachidic, eicosenoic fatty acids percent x 100

A 572 x 10 numeric array

# Importing Data

```{r}
# install.packages("cepp")
library(cepp)
data("olive")

write.csv(olive, file = "Olive Oil Data - Forina, Armanino, Lanteri, Tiscornia (1983).csv", na = "", row.names = FALSE)
```

# Data Pre-processing

```{r}
olive_df <- rename.vars(data = olive, from = colnames(olive), to = tools::toTitleCase(colnames(olive)), info = FALSE)

# Create vector for targets and features
target <- colnames(olive_df)[1:2]
features <- setdiff(colnames(olive_df), target)

# Rename values of Regions and Areas
olive_df$Region <- plyr::revalue(x = as.character(olive_df$Region), 
                                 replace = c("1" = "South", 
                                             "2" = "Sardinia", 
                                             "3" = "North"))

olive_df$Area <- plyr::revalue(x = as.character(str_trim(olive_df$Area)), 
                               replace = c("North-Apulia" = "Apulia (North)",
                                           "South-Apulia" = "Apulia (South)",
                                           "Inland-Sardinia" = "Sardinia (Inland)",
                                           "Coast-Sardinia" = "Sardinia (Coast)",
                                           "East-Liguria" = "Liguria (East)",
                                           "West-Liguria" = "Liguria (West)"))

# Convert features into %
olive_df[, features] <- olive_df[, features]/100
```

Some observations do not add up to 100%

```{r}
hist(rowSums(olive_df[, features]), main = "Histogram of Total of Fatty Acid Content", xlab = "Total (%)")
```

Adjust fatty acid content to add to 100% for all samples

```{r}
adjust_to_100 <- function(x) { (x/sum(x))*100 }

olive_df_adjusted <- olive_df

for (i in 1:nrow(olive_df_adjusted)) {
  
  olive_df_adjusted[i, features] <- adjust_to_100(olive_df[i, features])
  
}

# # Check adjustment was successful
# rowSums(olive_df_adjusted[,features])
# hist(rowSums(olive_df_adjusted[, features]), main = "Histogram of Total of Fatty Acid Content", xlab = "Total (%)")
```

# Exploratory Data Analysis

```{r}
# Hmisc::describe(olive_df_adjusted)

means_area <- olive_df_adjusted %>% 
  dplyr::group_by(Area) %>% 
  gather(key = "fatty_acid", value = "percentage", -Region, -Area) %>% 
  dplyr::group_by(Region, Area, fatty_acid) %>% 
  dplyr::summarise(Mean = mean(percentage, na.rm = TRUE)
                   # Std_dev = sd(percentage, na.rm = TRUE),
                   # Min = min(percentage, na.rm = TRUE),
                   # Max = max(percentage, na.rm = TRUE),
                   # Count = n()
  ) %>% 
  spread(key = fatty_acid, value = Mean)

# Print table of mean values
means_area %>% knitr::kable(digits = 2, align = "c")

means_fatty_acid <- apply(means_area[,features], MARGIN = 2, mean)
std_dev_fatty_acid <- apply(means_area[,features], MARGIN = 2, sd) 
  
means_area_scaled <- means_area

means_area_scaled[, features] <- scale(means_area_scaled[, features], center = means_fatty_acid, scale = std_dev_fatty_acid)

# # Check scaling output
# apply(means_area_scaled[, features], MARGIN = 2, sd)
# apply(means_area_scaled[, features], MARGIN = 2, mean) %>% round(8)

means_area_scaled %>% 
  gather(key = "fatty_acid", value = "percentage", -Area, -Region) %>% 
  dplyr::arrange(Region, Area) %>% 
  ggplot(aes(x=fatty_acid,  y=Area, fill=percentage)) +
    geom_tile() +
    labs(title = "Z-scores of mean fatty acid composition of olive samples by area", x = "Fatty Acid", y = "Area", fill = "z-score") +
    scale_fill_viridis_c() +
    theme_bw()
  
```

```{r fig.width=12, fig.height=6}
olive_df_adjusted %>% 
  dplyr::select(-Area) %>% 
  gather(key = "Fatty Acid", value = "Percentage", -Region) %>% 
  ggplot(mapping = aes(x = Percentage, fill = `Fatty Acid`)) +
    geom_histogram(bins = 30, alpha = 0.7) +
    facet_grid(Region~`Fatty Acid`, scales = "free_x") +
    scale_fill_brewer(palette = "Set1") +
    labs(title = "Distribution of fatty acids for each region",
         # subtitle = "Source: AFSIS",
         y = "Count",
         x = "%") +
    theme_bw() + theme(legend.position = "none")
```

# Clustering

## K-Means

Perform iterative k-means clustering for various k values and compute the within-cluster sums of squares for each value of k.

Each k value is tested with 20 random starts. The Hartigan-Wong algorithm ([Hartigan, J. A. and Wong, M. A. (1979)](https://www.jstor.org/stable/2346830?origin=crossref&seq=1)) will be used.

```{r}
olive_df_scaled <- olive_df_adjusted
olive_df_scaled[, features] <- scale(olive_df_scaled[, features], center = TRUE, scale = TRUE)

# # Check scaling was successfull
# apply(olive_df_scaled[, features], MARGIN = 2, mean) %>% round(8)
# apply(olive_df_scaled[, features], MARGIN = 2, sd)

wss <- NULL # vector for within sum of squares
bss <- NULL # vector for between sum of squares
set.seed(1) # set seed for reproducibility

for (i in 1:20) {
  
  kmeans_output <- kmeans(x = olive_df_scaled[, features], centers = i, nstart = 20, iter.max = 30, algorithm = "Hartigan-Wong")
  wss[i] <- kmeans_output$tot.withinss
  bss[i] <- kmeans_output$betweenss
  
}

kmeans_results <- data.frame(k = 1:20,
                             WSS = wss,
                             BSS = bss)

kmeans_results %>% 
  gather(key = "Parameter", value = "Value", -k) %>% 
  ggplot(aes(x = k, y = Value, col = Parameter)) +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = 1:20, minor_breaks = NULL) +
    scale_color_brewer(type = "qual", palette = "Set1") +
    labs(title = "Total Within Sum of Square and Total Between Sum of Square\nfor each k value tested") +
    theme_light()
```

### Selecting k

We will select a k of 5 for further k-means clustering on this dataset.

```{r}
set.seed(1)
kmeans_output <- kmeans(x = olive_df_scaled[, features], centers = 5, nstart = 20, iter.max = 30, algorithm = "Hartigan-Wong")

olive_df_scaled$kmeans_clusters <- kmeans_output$cluster
olive_df_adjusted$kmeans_clusters <- kmeans_output$cluster

olive_df_scaled <- olive_df_scaled %>% 
  dplyr::mutate(kmeans_clusters = kmeans_output$cluster) %>% 
  dplyr::select(Region, Area, kmeans_clusters, everything())

```

### Confusion Matrix

```{r}
confusion_matrix_region <- olive_df_scaled %>% 
   dplyr::select(Region, Area, kmeans_clusters) %>% 
   dplyr::mutate(idx = 1) %>% 
   dplyr::group_by(Region, Area, kmeans_clusters) %>%
   dplyr::summarise(Frequency = sum(idx)) %>% 
   spread(key = "kmeans_clusters", value = "Frequency", fill = 0) %>% 
   dplyr::arrange(Region, Area)
```


```{r}
confusion_matrix_region %>% 
  # ungroup(Region) %>%
  gather(key = "kmeans_clusters", value = "Frequency", -Area, -Region) %>% 
  dplyr::mutate(kmeans_clusters = as.numeric(kmeans_clusters)) %>% 
  ggplot(aes(x = kmeans_clusters, y = fct_reorder(Area, dplyr::desc(Region)), fill = Frequency, label = Frequency)) + 
  geom_tile() +
  geom_text() +
  scale_fill_gradient(low = "#f0f0f0", high = "#525252", guide = "none") +
  labs(x = "K-means Clusters", y = "", title = "Confusion matrix of k-means clusters and area of olive oil samples") +
  theme_light()
```


```{r}
table(olive_df_scaled %>% dplyr::select(Region, kmeans_clusters)) %>% 
  data.frame() %>% 
  ggplot(aes(x = kmeans_clusters, y = fct_reorder(Region, desc(Region)), fill = Freq, label = Freq)) + 
  geom_tile() +
  geom_text() +
  scale_fill_gradient(low = "#f0f0f0", high = "#525252", guide = "none") +
  labs(x = "K-means Clusters", y = "", title = "Confusion matrix of k-means clusters and region of olive oil samples") +
  theme_light()
```

## Gaussian Mixture Models

```{r}
gmm <- mclust::Mclust(data = olive_df_adjusted[, features], G = 2:10, verbose = FALSE)

summary(gmm)
```

```{r}
olive_df_adjusted$gmm_clusters <- gmm$classification
olive_df_adjusted$gmm_uncertainty <- gmm$uncertainty

gmm_probability <- data.frame(gmm$z)
colnames(gmm_probability) <- paste0("GMM Cluster ", str_pad(1:gmm$G, width = 2, side = "left", pad = "0"))
olive_df_adjusted <- cbind(olive_df_adjusted, gmm_probability)
```


```{r}
gmm_BIC <- matrix(gmm$BIC, nrow = dim(gmm$BIC)[1], ncol = dim(gmm$BIC)[2]) %>% data.frame() %>% add_column(K = 2:10, .before=1)
colnames(gmm_BIC) <- c("K", colnames(gmm$BIC))

gmm_BIC %>% 
  gather(key = "Model", value = "BIC", -K) %>% 
  filter(!is.na(BIC)) %>% 
  ggplot(aes(x = K, y = BIC, col = Model)) +
    geom_line(size=0.75) +
    geom_point() +
    scale_color_brewer(type = "qual", palette = "Set1", direction = -1) +
    scale_x_continuous(breaks = 1:10, minor_breaks = NULL) +
    labs(title = "Gaussian Mixture Models",
         subtitle = "BIC scores for components ranging from 1 to 10 and various covariance models",
         x = "Number of Components") +
    theme_light()
```


```{r fig.align='center', fig.height=5, fig.width=12}
olive_df_adjusted %>% 
  dplyr::select(Region, Area, contains("GMM Cluster")) %>% 
  mutate(ID = row_number()) %>% 
  gather(Cluster, probability, -ID, -Region, -Area) %>% 
  dplyr::arrange(Region, Area) %>% 
  ggplot(aes(x = probability, fill = Area)) +
    geom_histogram() +
    facet_wrap(~ Cluster, nrow = 2) +
    labs(x = "Probability", y = "Count",
         title = "Distribution of probabilities for all observations aligning to each cluster") +
    scale_fill_brewer(type = "qual", palette = "Set1") + 
    theme_light()
```







































